---
title: "20261008_visualization"
author: "Hongtong Lin"
date: "2026-10-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# loading libraries
library(tidyverse)
library(patchwork) # to allow two ggplots to be included in one window
library(p8105.datasets)
library(viridis)
data("weather_df")

# color palette setting
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
```



```{r}
weather_df |> 
  filter(prcp >0) |> 
  ggplot(aes(x = prcp)) + 
  geom_histogram()
```
  
Do I believe this value?  
  
```{r}
weather_df |> 
  ggplot(aes(x = tmax, y = tmin, color = name, shape = name)) + 
  geom_point()
```


## group_by()  

add some groups  

```{r}
weather_df <- 
  weather_df |> 
  mutate(
    month = lubridate::floor_date(date, unit = "month")
  ) # lowest value for each of the month

weather_df
```

```{r}
weather_df |> 
  group_by(name, month) # 3 names, 12 months, 2 years, multipled together = 72 groups
```
  
  
## `summarize()`  

```{r}
weather_df |> 
  group_by(name) |> 
  summarise(count = n())
```

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(count = n())
```

```{r}
weather_df |> 
  count(name) # instead of doing group_by and summarizing, it will automatically counting for you. 
```

```{r}
weather_df |> 
  group_by(month) |> 
  summarise(
    count = n(),
    n_days = n_distinct(date),
    )
```
 
  
```{r}
weather_df |> 
  group_by(name) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmax = median(tmax, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE),
    q25_prcp = quantile(prcp, .25, na.rm = TRUE),
    q95_prcp = quantile(prcp, .95, na.rm = TRUE)
  ) |> 
  knitr::kable(digits = 2) # reader friendly, limit the signficance figures to be 2at max
```

```{r}
weather_df |> 
  group_by(name) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) # can make the table untidy to be reader friendly
  
```
  
  
what about doing this  

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  ggplot(aes(x =month, y= mean_tmax, color = name)) +
  geom_point() + 
  geom_line()
```

```{r}
weather_df |> 
  mutate(
    center_tmax = tmax - mean(tmax, na.rm = TRUE)
  ) |> 
ggplot(aes(x= date, y = center_tmax, color = name))+
  geom_point()
```

What about "window" functions?  

Try to rank things  

```{r}
weather_df |> 
  group_by(name, month) |> 
  mutate(temp_rank = min_rank(tmax)) |> 
  filter(temp_rank <2)
```

`lead` and `lag` functions  
the order of the data sets matters
```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    lagged_tmax = lag(tmax) # Shift by 1 position (default)
    )
```

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    lagged_tmax = lag(tmax, n = 3) # Shift by 3 positions
    )
```

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    lagged_tmax = lag(tmax), # Shift by 3 positions
    temp_change = tmax - lagged_tmax
    ) |> 
  summarise(
    mean_temp_change = mean(temp_change, na.rm = TRUE),
    sd_temp_change = sd(temp_change, na.rm = TRUE)
  )
```


## revisit same samples  

import, clean, tidy, etc the pulse data, and compute mean and median BDI score at each visit  

```{r}
pulse_df <- haven::read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    values_to = "bdi_score"
  ) |> 
  select(id, visit, everything()) |> 
  mutate(
    visit = replace(visit, visit == "bl", "00m")
  ) 
# |> 
#   str_remove(visit, "bdi_score_")

pulse_df |> 
  group_by(visit) |> 
  summarize(
    mean_bdi = mean(bdi_score, na.rm=TRUE),
    median_bdi = median(bdi_score, na.rm=TRUE)
    )
```


In the FAS data, compute mean outcome (ears only) across dose and day_of_txl; show in a reader friendly table  

```{r}
pups_df <- 
  read_csv("data/FAS_pups.csv", skip = 3, na = c("NA", "", ".")) |> 
  janitor::clean_names() 

litter_df <-
  read_csv("data/FAS_litters.csv", na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  separate(group, into = c("dose", "day_of_tx", 3))

fas_df <- 
  left_join(pups_df, litter_df, by = "litter_number") |> 
  select(litter_number, dose, day_of_tx, everything())

fas_df |> 
  group_by(dose, day_of_tx) |> 
  summarise(
    mean_ears = mean(pd_ears, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = day_of_tx, 
    values_from = mean_ears
  )
```




